Disparities in Climate-Induced Health Outcomes in the Greater Toronto Area.

### 2022-07-27 outputs/paper/paper.qmd
Updates to paper

Carbon monoxide is a product of incomplete combustion of hydrocarbon-based fuels. In figure 12 and 13 we see a similar pattern as NO and NO2 where the level of pollution is decreasing throughout the years from 2002 to 2020 and the highest levels are in the winter months. The highest level of pollution for CO was seen in 2002 with 1762.5 ppb. This level is much higher than any of the other pollutants. This suggests that carbon monoxide levels are much more widespread than any of the other pollutants we have analyzed and due to this can have a affect on our health. Even with the decreasing levels, the affects will still be dangerous. The highest level in 2020 was around 550 ppb which is almost 5 times higher than the levels of NO, NO2, and O3 in 2020. There is also a decreasing pattern for the mean pollution from year 2002 to 2020 for CO by year. We can notice that there is a drop in the pollution levels from 2003 to 2005 but it rises again in 2006. Looking back at figure 3, this was the same year some of the highest temperatures were recorded from August to November. The highest levels of CO are during winter and the colder months. This might be due to the fact that heaters are used the most in this time and do emit carbon monoxide. This plot does not tell us much else about the other pollutants due so from here we will take a look at each pollutant and how it has behaved from 2002 to 2020 by year and month.

@@

NO2 is usually formed through the burning of fossil fuels, transportation services, energy production and industrial processes [@citeairquality]. Nitrogen dioxide at higher concentration levels has a strong and harsh odor and can be seen over large cities as a brownish haze [@citeairquality]. NO2 can combine with H2O and form nitric acid and nitrous acid. This falls to earth through precipitation like rain, snow, and fog, and contributed to acidification and eutrophication of ecosystems [@citeairquality].

SO2 emissions come from natural sources like volcanoes and forest fires but these emissions are small compared to those from industrial sources [@citeairquality]. Environment Canada accounted for 1 million tonnes of man-made SO2 emissions compared to 100 tonnes of natural SO2 emissions in 2022. SO2 is seen in acid rain and due to this early recognition policies were put in place that have reduced SO2 pollution in recent decades as seen in figure 14.

@@

Table 3 shows

@@

Table 4 shows ...

Exposure to air pollution is known as a major cause to illnesses and death.

SO2 (Sulfur Dioxide) can harm human respiratory systems by triggering respiratory illness and worsening existing respiratory and cardiovascular diseases [@citepembinainstitute]. Sulfur dioxide is seen in smog and can react with other substances like nitric oxide to harm human health [@citeconferenceboard]. When SO2 transforms into sulfate particles and combines with other particles in the atmosphere it can contribute to the formation of particulate matter (PM2.5) which harms human health as well as the environment [@citeconferenceboard]. Sensitive populations such as children, the elderly, those with asthma are more susceptible to the affects of SO2 [@citepembinainstitute].

NO2 (Nitrogen Dioxide) has consequences on human health including emergency hospitalizations for respiratory conditions for adults that live in lower income communities in Toronto \[\@citestatscan\]. Children are more vulnerable to these effects of respiratory damage \[\@citestatscan\]. Children that have been exposed to NO2 have an increased risk of asthma, wheezing, infections in the ear, nose and throat, influenza, serious colds, and reduced lung function \[\@citestatscan\].

>>>>

Carbon monoxide is produced through burning of coal, gasoline, natural gas, oil, propane, and wood [@citeCOpoisoning]. In figure 12 and 13 we see a similar pattern as NO and NO2 where the level of pollution is decreasing throughout the years from 2002 to 2020 and the highest levels are in the winter months. The highest level of pollution for CO was seen in 2002 with 1762.5 ppb. This level is much higher than any of the other pollutants. This suggests that carbon monoxide levels are much more widespread than any of the other pollutants we have analyzed and due to this can have a affect on our health. Even with the decreasing levels, the affects will still be dangerous. The highest level in 2020 was around 550 ppb which is almost 5 times higher than the levels of NO, NO2, and O3 in 2020. There is also a decreasing pattern for the mean pollution from year 2002 to 2020 for CO by year. We can notice that there is a drop in the pollution levels from 2003 to 2005 but it rises again in 2006. Looking back at figure 3, this was the same year some of the highest temperatures were recorded from August to November. The highest levels of CO are during winter and the colder months. This might be due to the fact that heaters are used the most in this time and do emit carbon monoxide. This plot does not tell us much else about the other pollutants due so from here we will take a look at each pollutant and how it has behaved from 2002 to 2020 by year and month.

@@

NO2 is usually formed through the burning of fossil fuels, transportation services, energy production and industrial processes [@citeairquality]. Nitrogen dioxide at higher concentration levels has a strong and harsh odor and can be seen over large cities as a brownish haze [@citeairquality]. NO2 can combine with H2O and form nitric acid and nitrous acid. This falls to earth through precipitation like rain, snow, and fog, and contributed to acidification and eutrophication of ecosystems [@citeairquality]. NO2 (Nitrogen Dioxide) has consequences on human health including emergency hospitalizations for respiratory conditions for adults that live in lower income communities in Toronto \[\@citestatscan\]. Children are more vulnerable to these effects of respiratory damage \[\@citestatscan\]. Children that have been exposed to NO2 have an increased risk of asthma, wheezing, infections in the ear, nose and throat, influenza, serious colds, and reduced lung function \[\@citestatscan\].

SO2 emissions come from natural sources like volcanoes and forest fires but these emissions are small compared to those from industrial sources [@citeairquality]. Environment Canada accounted for 1 million tonnes of man-made SO2 emissions compared to 100 tonnes of natural SO2 emissions in 2022. SO2 is seen in acid rain and due to this early recognition policies were put in place that have reduced SO2 pollution in recent decades as seen in figure 14. SO2 (Sulfur Dioxide) can harm human respiratory systems by triggering respiratory illness and worsening existing respiratory and cardiovascular diseases [@citepembinainstitute]. Sulfur dioxide is seen in smog and can react with other substances like nitric oxide to harm human health [@citeconferenceboard]. When SO2 transforms into sulfate particles and combines with other particles in the atmosphere it can contribute to the formation of particulate matter (PM2.5) which harms human health as well as the environment [@citeconferenceboard]. Sensitive populations such as children, the elderly, those with asthma are more susceptible to the affects of SO2 [@citepembinainstitute].

@@

Table 3 shows the mean pollution levels of each pollutant from 2002 to 2020. The highest mean is from CO levels followed by O3. From the previous analysis, carbon monoxide is produced through burning of coal, gasoline, natural gas, oil, propane, and wood and as these activities have been continuing throughout the city of Toronto thus CO levels remain high [@citeCOpoisoning]. O3 levels have also been increasing from 2002 to 2020 while the remaining pollutants' levels (NO, NO2, and SO2) have been decreasing in that time.

@@

Table 4 shows the mean pollution levels of each pollutant by five-year intervals. From this table, it is evident that NO, NO2, SO2, and CO levels have been decreasing since 2002 while O3 has been increasing. The levels of CO still remain higher than any of the other pollutants. The biggest difference in pollution levels are seen for NO. From 2002 to 2006 the mean was 15.05 but this dropped to 3.25 from 2017 to 2020.
